---
title: "Citibike Station Use"
author: "Vitaly Druker"
date: "November 20th, 2017"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
# to enable presenter mode add ?presentme=true to the URL of the presentation

knitr::opts_chunk$set(echo = FALSE,
                      comment = NA,
                      warning = FALSE,
                      # cache = TRUE,
                      fig.align = 'center', 
                      # fig.height = 6,
                      dev.args = list(bg = 'transparent'))

library(ggplot2)
library(dplyr)
library(lubridate)
library(knitr)
library(pander)
library(tibble)
options(tibble.print_max = 5)

library(readr)
library(tidyr)
library(ggmap)
library(stringr)
library(lme4)

library(lattice)
library(sp)
library(gstat)

source('../stat 787 - hw/helper_functions.R')

theme_set(theme_minimal())
theme_update(panel.grid.minor = element_blank())
theme_update(plot.background = element_rect(fill = "transparent", color = NA))
theme_update(legend.position = "bottom")
```

```{r load_data}
load('transformed_data/citibike_rides.Rdata')

labeled_stations <-
  read_csv(
    'transformed_data/labeled_stations.csv',
    col_types = cols(
        boro_name = col_character(),
  boro_code = col_integer(),
  shape_leng = col_double(),
  shape_area = col_double(),
      startstationid = col_character(),
      startstationname = col_character(),
      startstationlatitude = col_double(),
      startstationlongitude = col_double(),
      cluster = col_character()
    )
  )

```


## Overview

1. Problem Overview
2. What Data is Used?
3. Regression Analysis
4. Spatial Analysis

# Problem Overview

## Question 1

Does the proximity of opening a new station effect the ridership of other stations near the newly opened station?

> - Not able to turn this into a 'spatial data' problem, seems like a distance problem

## Question 2 

How does the proximity of NYC subway stations effect ridership?

> - Bike Share guide recommends placing bike stations near public transit. [NACTO](https://nacto.org/wp-content/uploads/2016/04/NACTO-Bike-Share-Siting-Guide_FINAL.pdf)

# Data Sources 

## Citibike Data | Source {.build}

[Ridership Data](https://www.citibikenyc.com/system-data)

```{r}
d_filt %>% 
  select(-startstationlatitude, -startstationlongitude, 
         -endstationlatitude, -endstationlongitude,
         -mnth_trip, -day_trip) %>% 
  glimpse
```

## Citibike Data | Summary Statistics {.build}

Time: 6 - 10 am weekday mornings

Date Range: 
```{r}
range(d_filt$day_trip)
```

Stations:
```{r}
nrow(labeled_stations)
```

## Citibike Data | Exploration {.build}

<div class="columns-2">

```{r explore-citi, out.width="100%", fig.height=6, fig.width = 6}

station_cts <- d_filt %>% 
  select(mnth_trip, startstationid) %>% 
  left_join(labeled_stations, by = "startstationid") %>% 
  group_by(mnth_trip,boro_name) %>% 
  summarise(stations = n_distinct(startstationid),
            rides = n())

station_cts %>% 
  ggplot(aes(x = mnth_trip, y = rides, color = boro_name)) + geom_line(stat = "identity") +
  labs( title = "Monthly Citibike Rides",
        x = "Month",
        y = "Number of Rides",
        color = "") +
  ylim(c(0,NA))

```

```{r explore-ctit-3, out.width="100%", fig.height=6, fig.width = 6}

station_cts %>% 
  ggplot(aes(x = mnth_trip, y = stations, fill = boro_name)) + geom_bar(stat = "identity") +
  labs( title = "Total Citibike Stations",
        x = "Month",
        y = "Number of Stations",
        fill = "")

```

</div>


## Citibike Data | Exploration (Cont.) {.build}

```{r, cache=TRUE}

daily_cts <- d_filt %>% 
  count(day_trip, startstationid) 


sbbox <- with(labeled_stations, 
              make_bbox(lon = startstationlongitude, lat = startstationlatitude, f = .1)
)

suppressMessages(sq_map <- get_map(sbbox,source = 'stamen', maptype = "terrain-lines", messaging = FALSE))



samp_days <- as.POSIXct(c("2015-07-01", "2016-01-04", "2016-07-01","2017-07-03"))
  

mapping_data <- daily_cts %>% 
  spread(startstationid, n, fill = NA) %>% 
  gather("startstationid","n", -day_trip) %>% 
  left_join(labeled_stations, by = "startstationid") %>% 
  filter(day_trip %in% samp_days) %>% 
  filter(startstationlongitude != 0) %>% 
  filter(!is.na(n)) %>% 
  mutate(state = "new york")

ggmap(sq_map) +
  geom_point(aes(x = startstationlongitude, y = startstationlatitude, size = n, color = boro_name), 
             alpha = 0.3, data = mapping_data) + facet_wrap(~day_trip, nrow = 1) +
  theme_inset() + labs(title = "Daily Ridership Through the Years")
```


## Citibike Data {.build}

<div class="columns-2">

```{r, fig.width=3, fig.height = 6}
ggmap(sq_map) +
  geom_point(aes(x = startstationlongitude, y = startstationlatitude, color = cluster), 
             data = labeled_stations) +
  theme_inset()+
labs(title= "Clustered Stations")
  
```

- Only  the stations labeled in red will be used in the analysis.

- These stations have the largest amount of longitudinal data and are more homogenous in the surrounding area.

- Start analysis in 2014, giving 6 months "burn in"

```{r, cache = TRUE}
labeled_stations <- labeled_stations %>% filter(cluster == 1)


labeled_stations <- labeled_stations %>% 
  distinct(startstationid, .keep_all = TRUE)


stat_clean <- labeled_stations %>%
  select(id = startstationid,
         lat = startstationlatitude,
         lon = startstationlongitude) %>% 
  mutate(k = 1)

stat_clean <- inner_join(stat_clean, stat_clean, by = "k") %>% 
  select(-k)

station_distance <- stat_clean %>% 
  mutate(dist = gcd_slc(lon.x, lat.x, lon.y, lat.y)) %>% 
  filter(id.x!=id.y)

rm(stat_clean)

d_filt <- d_filt %>% 
  semi_join(labeled_stations, by = "startstationid") %>% 
  filter(day_trip >= '2014-01-01')


sbbox <- with(labeled_stations, 
              make_bbox(lon = startstationlongitude, lat = startstationlatitude, f = .1)
)

suppressMessages(sq_map <- get_map(sbbox,source = 'stamen', maptype = "terrain-lines", messaging = FALSE))


```

</div>

## Historic Weather Data

### [NOAA Data Request](https://www.ncdc.noaa.gov/cdo-web/)

```{r}

weather_data <- read_csv(
  'weather_data/NYC_20130101-20170731.csv',
  col_types = cols(
    .default = col_integer(),
    STATION = col_character(),
    NAME = col_character(),
    DATE = col_character(),
    AWND = col_double(),
    PGTM = col_character(),
    PRCP = col_double(),
    SNOW = col_double(),
    TAVG = col_character(),
    WSF2 = col_double(),
    WSF5 = col_double()
  )
)
weather_data <- weather_data %>%
  mutate(day_trip = mdy(DATE),
         day_trip = as.POSIXct(day_trip)) %>%
  select(day_trip, PRCP, SNOW, SNWD, TMIN, TMAX)

weather_data %>% glimpse
```

## Subway Entrance Data 

[Subway Entrance Data](https://data.ny.gov/Transportation/NYC-Transit-Subway-Entrance-And-Exit-Data/i9wp-a4ja)

```{r}
subway_data <-
  read_csv(
    'subway_data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv',
    col_types = cols(
      .default = col_character(),
      `Station Latitude` = col_double(),
      `Station Longitude` = col_double(),
      Route8 = col_integer(),
      Route9 = col_integer(),
      Route10 = col_integer(),
      Route11 = col_integer(),
      ADA = col_logical(),
      `Free Crossover` = col_logical(),
      `Entrance Latitude` = col_double(),
      `Entrance Longitude` = col_double()
    )
    
  )

subway_data <- subway_data %>% 
  select(subway_name = `Station Name`, 
         subway_lat = `Station Latitude`,
         subway_long = `Station Longitude`) %>% 
  distinct()

subway_data2 <- labeled_stations %>% 
  mutate(k = 1) %>% 
  inner_join(subway_data %>% mutate(k = 1), by = "k") %>% 
  mutate(dist = gcd_slc(startstationlongitude, startstationlatitude, 
                        subway_long, subway_lat)) %>% 
  group_by(startstationid) %>% 
  summarise(closest_subway = min(dist),
            close_subways = sum(dist <=0.5))
 
labeled_stations <- labeled_stations %>% 
  left_join(subway_data2,by = "startstationid")
 
```


<div class="columns-2">

```{r explore-subway, out.width="100%", fig.height=6, fig.width = 6}

labeled_stations %>% 
  ggplot(aes(x = closest_subway)) + geom_histogram(binwidth = 0.05) +
  labs(title = "Closest Subway to Citibike (KM)",
       x = "Distance to Closest Subway")

```

```{r explore-subway-2, out.width="100%", fig.height=6, fig.width = 6}

labeled_stations %>% 
  ggplot(aes(x = close_subways)) + geom_histogram(binwidth = 1) +
  labs(title = "Subway Stations within 0.5 KM",
       x = "Number of Close Subway Stations")

```

</div>
# Regression Analysis


## Model Definition

- Fit a hierarchical Poisson model to control for variability throughout the year.

> -  Outcome - Number of Rides 

### Coefficients:

> - Fixed: Temperature, Snow, Days from Jan 1, 2014, previous Days count
> - Random: Station, Week of Year

```{r}
regression_data <- d_filt %>% 
  count(day_trip,startstationid) %>% 
  left_join(weather_data, by = "day_trip") %>% 
  mutate(days = day_trip-min(day_trip),
         days = as.numeric(days, unit = "days")) %>% 
  mutate(week_num = week(day_trip))
```


```{r, cache = TRUE}
basic_mod <- glmer(n ~ scale(PRCP) + scale(SNOW) + scale(SNWD) + #TMIN + 
                     scale(log(TMAX)) + (1|startstationid) +(1|week_num)+scale(log(days + 1)) + scale(lag(n))
                   # + cos(2*pi*days) + sin(2 * pi*days) 
                     + I(cos(2*pi*days/365)) + I(sin(2 * pi*days/365))
                   # + cos(2*pi*days/30) + sin(2 * pi*days/30)
                   , 
                   data = regression_data, family = poisson)

int_only_mod <-glmer(n ~ (1|startstationid), 
                   data = regression_data, family = poisson)

regression_data <- regression_data %>% 
  mutate(full_fit = predict(basic_mod, ., type = "response"),
         simple_fit = predict(int_only_mod, ., type = "response"),
         full_resid = n - full_fit)
```

## Model Performance 

```{r}
regression_data %>% 
  filter(startstationid %in% c('151')) %>% 
  ggplot(aes(x = day_trip)) + 
    geom_smooth(aes(y = n - simple_fit, color = "Average"), alpha = 0.5, method = "loess", span = 0.2)+
  geom_smooth(aes( y = n-full_fit, color = "Full Regression"), alpha = 0.5, method = "loess", span = 0.2)  +
  scale_color_manual(
    name = "",
    values = c("#999999","#D55E00"),
    labels = c("Average", "Full Regression")
  ) +
  labs(y = "Smoothed Residuals",
       x = "Date",
       title = "Regression Model Performance")

```


## Specific Group of Stations

```{r}
regression_data %>% 
  filter(startstationid %in% c('3427','251','151')) %>% 
  filter(day_trip >= '2015-01-01') %>% 
  ggplot(aes(x = day_trip, y = full_resid, color = startstationid)) +
    geom_smooth(span = 0.2, method = "loess") +
  labs(title = "New Station's Effect on Surrounding Stations",
       y = "Smoothed Residuals",
       x = "Day of Trip",
       color = "Station ID")
```


# Spatial Analysis

## Models

Does proximity to subway stations show decrease in Citibike usage?

```{r}
spatial_data <- d_filt %>% 
  count(day_trip, startstationid) %>% 
  inner_join(labeled_stations, by = "startstationid")

single_day <- spatial_data %>% 
  filter(day_trip =='2016-07-01') %>% 
  filter( n < quantile(n, .99))

ggmap(sq_map) + 
  geom_point(aes(x = subway_long, y= subway_lat), data = subway_data) + 
  geom_point(aes(x = startstationlongitude, y = startstationlatitude, color = n, size = n), 
             alpha = 0.7, data = single_day) +
  theme_inset() +
  labs(title = "Citibike Ridership on July 1st, 2016,\nwith subway stations overlaid.")

coordinates(single_day) <- c("startstationlongitude", "startstationlatitude")
```

## Fitting the Variogram
<div class = "columns-2">

### No Regressors

```{r, out.width="100%", fig.height=5, fig.width = 5}
vgm_sel <- vgm(.3, c("Sph"), nugget = 0.01)

vt <- variogram(log(n+1) ~ 1, single_day, cutoff = .020, width = .020/15)
vfit <- (fit.variogram(vt, vgm_sel))
plot(vt, vfit)
```

### Closest Subway Features

```{r, out.width="100%", fig.height=5, fig.width = 5}
vt2 <- variogram(log(n+1) ~ sqrt(closest_subway) + log(close_subways + 1), single_day, cutoff = 0.020, width = .020/15)
vfit2 <- (fit.variogram(vt2, vgm_sel))
plot(vt2, vfit2)

```
</div>

# Next Steps

## Further Analysis

### Regression

> - Group a month before and a month after and compare the difference, model that as distance from new station.

### Spatial Analysis

> - How can we tell if a predictor is 'significant'?


<!-- ```{r} -->
<!-- library(purrr) -->
<!-- set.seed(1) -->
<!-- out <- spatial_data %>%  -->
<!--   left_join(regression_data %>% select(day_trip, startstationid, full_resid)) %>%  -->
<!--   select(-n) %>%  -->
<!--   rename(n = full_resid) %>%  -->
<!--   filter(!is.na(n)) %>%  -->
<!--   filter(day_trip %in% sample(day_trip, 10)) %>%  -->
<!--   # filter(n<200) %>%  -->
<!--    nest(-day_trip) %>%  -->
<!--   mutate(data = map(data, function(x){ -->
<!--     coordinates(x) <- c("startstationlongitude", "startstationlatitude") -->
<!--     x}), -->
<!--     vt = map(data, ~variogram(n ~ sqrt(closest_subway), .x, cutoff = 1/100, width = .5/100/15)), -->
<!--     vfit = map(vt, ~fit.variogram(.x, vgm(.3, c("Exp"), nugget = .01))), -->
<!--     krige_cv = map2(data, vfit, ~krige.cv(n ~sqrt(closest_subway), .x, .y, nfold = 5, verbose = FALSE)), -->
<!--     df = map(vfit, as.data.frame))  -->

<!-- out$krige_cv[[1]]$zscore %>%  -->
<!--   var() -->




<!-- bind_rows( -->
<!-- out %>% select(day_trip, model, val = psill) %>% mutate(var = "psill"), -->
<!-- out %>% select(day_trip, model, val = range) %>% mutate(var = "range") -->
<!-- ) %>%  -->
<!--   group_by(model, var) %>%  -->
<!--   summarise(mean(val), -->
<!--             median(val), -->
<!--             low = quantile(val,.05), -->
<!--             high = quantile(val, .95)) -->


<!-- bind_rows( -->
<!-- out %>% select(day_trip, model, val = psill) %>% mutate(var = "psill"), -->
<!-- out %>% select(day_trip, model, val = range) %>% mutate(var = "range") -->
<!-- )  -->


<!-- krige(n ~ sqrt(closest_subway), out$data[[1]]) -->


<!-- out  %>%  -->
<!--   ggplot(aes(psill, range)) + geom_point() + facet_wrap(~model) -->


<!--   ggplot(aes(x = val)) + geom_histogram() + facet_grid(model ~ var, scales = "free") + scale_x_log10() -->

<!--   group_by(model, var) %>%  -->
<!--   summarise(mean(val), -->
<!--             median(val), -->
<!--             low = quantile(val,.05), -->
<!--             high = quantile(val, .95)) -->

<!-- library(modelr) -->

<!-- spatial_data %>%  -->
<!--   left_join(regression_data %>% select(day_trip, startstationid, full_resid)) %>%  -->
<!--   select(-n) %>%  -->
<!--   rename(n = full_resid) %>%  -->
<!--   filter(!is.na(n)) %>%  -->
<!--   nest(-day_trip) %>%  -->
<!--   modelr::crossv_kfold() -->
<!--   # filter(day_trip %in% sample(day_trip, 10)) %>%  -->


<!--   nest() %>% mutate(data = map(data, function(x){ -->
<!--     coordinates(x) <- c("startstationlongitude", "startstationlatitude") -->
<!--     x}), -->
<!--     vt = map(data, ~variogram(n ~ sqrt(closest_subway), .x, cutoff = 1/100, width = .5/100/15)), -->
<!--     vfit = map(vt, ~fit.variogram(.x, vgm(.3, c("Exp"), nugget = .01))), -->
<!--     df = map(vfit, as.data.frame)) %>%  -->
<!--   unnest(df)  -->

<!-- ``` -->



