---
title: "Citibike Station Use"
author: "Vitaly Druker"
date: "November 20th, 2017"
output: 
  ioslides_presentation:
    widescreen: true
---

```{r setup, include=FALSE}
# to enable presenter mode add ?presentme=true to the URL of the presentation

knitr::opts_chunk$set(echo = FALSE,
                      comment = NA,
                      warning = FALSE,
                      # cache = TRUE,
                      fig.align = 'center', 
                      dev.args = list(bg = 'transparent'))

library(ggplot2)
library(dplyr)
library(lubridate)
library(knitr)
library(pander)
library(tibble)
options(tibble.print_max = 5)

library(readr)
library(tidyr)
library(ggmap)
library(stringr)
library(lme4)

source('../stat 787 - hw/helper_functions.R')


theme_update(plot.background = element_rect(fill = "transparent", color = NA))

```

```{r load_data}
load('transformed_data/citibike_rides.Rdata')

labeled_stations <-
  read_csv(
    'transformed_data/labeled_stations.csv',
    col_types = cols(
      startstationid = col_character(),
      startstationname = col_character(),
      startstationlatitude = col_double(),
      startstationlongitude = col_double(),
      cluster = col_character()
    )
  )

```

## Overview

1. Problem Overview
2. What Data is Used?
3. Regression Analysis
4. Spatial Analysis

# Problem Overview

## Question 1

Does the proximity of opening a new station effect the ridership of other stations near the newly opened station?

> - Not able to turn this into a 'spatial data' problem, seems like a distance problem

## Question 2 

How does the proximity of NYC subway stations effect ridership?

> - Bike Share guide recommends placing bike stations near public transit. [NACTO](https://nacto.org/wp-content/uploads/2016/04/NACTO-Bike-Share-Siting-Guide_FINAL.pdf)

# Data Sources 

## Citibike Data | Source {.build}

[Ridership Data](https://www.citibikenyc.com/system-data)

```{r}
d_filt %>% 
  select(-startstationlatitude, -startstationlongitude, 
         -endstationlatitude, -endstationlongitude,
         -mnth_trip, -day_trip) %>% 
  glimpse
```

## Citibike Data | Summary Statistics {.build}

Time: 6 - 10 am weekday mornings

Date Range: 
```{r}
range(d_filt$day_trip)
```

Stations:
```{r}
nrow(labeled_stations)
```

## Citibike Data | Summary Statistics {.build}





## Historic Weather Data

### [NOAA Data Request](https://www.ncdc.noaa.gov/cdo-web/)





